---
title: "Visualizing Plastics: Waste Composition and Environmental Cost"
author: "Nathan Lee, Noah Nayve, Ryan Ekdahl, Taiga Sobajima"
format:
  dashboard:
    orientation: rows
    theme: flatly
    logo: "logo.png"
    nav-buttons:
      - icon: envelope
        href: mailto:nlee174@calpoly.edu
        aria-label: Nathan Lee
      - icon: envelope
        href: mailto:nnayve@calpoly.edu
        aria-label: Noah Nayve
      - icon: envelope
        href: mailto:rekdahl@calpoly.edu
        aria-label: Ryan Ekdahl
      - icon: envelope
        href: mailto:tsobajim@calpoly.edu
        aria-label: Taiga Sobajima
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(glue)
library(gt)
library(plotly)
library(scales)
library(ggridges)
library(broom)
library(ggrepel)
```

```{r}
#| label: load-data
#| message: false
plastics <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
population_plastics <- read_csv("data/population_plastics.csv")
```

```{r}
#| label: prep
#| message: false
#| warning: false

summarize_plastic_and_ecology <- function(country_name, data = population_plastics) {
  if (!(country_name %in% data$country)) stop(paste("Country not found:", country_name))
  country_stats <- data |>
    filter(country == country_name, parent_company == "Grand Total", grand_total > 0) |>
    summarize(
      country = country_name,
      avg_pet = mean(pet / grand_total, na.rm = TRUE),
      avg_hdpe = mean(hdpe / grand_total, na.rm = TRUE),
      avg_ldpe = mean(ldpe / grand_total, na.rm = TRUE),
      avg_pp = mean(pp / grand_total, na.rm = TRUE),
      avg_ps = mean(ps / grand_total, na.rm = TRUE),
      avg_pvc = mean(pvc / grand_total, na.rm = TRUE),
      avg_o = mean(o / grand_total, na.rm = TRUE),
      avg_forested = mean(forested_area, na.rm = TRUE),
      avg_threatened = mean(threatened_species, na.rm = TRUE)
    )
  leading_plastics <- country_stats |>
    pivot_longer(cols = !c(country, avg_forested, avg_threatened),
                 names_to = "plastic_type", values_to = "avg_value") |>
    slice_max(order_by = avg_value, n = 1, with_ties = FALSE) |>
    mutate(plastic_type = str_remove(plastic_type, "avg_"),
           plastic_type = str_to_upper(plastic_type),
           plastic_type = if_else(plastic_type == "O", "Other", plastic_type)) |>
    pull(plastic_type)
  country_stats |> mutate(leading_plastic = leading_plastics)
}

# precompute all needed data
all_countries_rq2 <- population_plastics |>
  filter(parent_company == "Grand Total", grand_total > 0,
         !is.na(forested_area), !is.na(threatened_species)) |>
  distinct(country) |> pull(country)

plastic_ecology_summary_table <- map(all_countries_rq2, summarize_plastic_and_ecology) |>
  bind_rows()

top3_plastics <- population_plastics |>
  filter(parent_company == "Grand Total") |>
  summarize(across(c(pet, hdpe, ldpe, pp, ps, pvc), ~ sum(.x, na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "plastic", values_to = "total") |>
  slice_max(total, n = 3) |> pull(plastic)

gt_style <- function(gt_obj) {
  gt_obj |>
    tab_style(style = cell_fill(color = "#457B9D"), locations = cells_column_labels()) |>
    tab_style(style = cell_text(color = "white", weight = "bold"), locations = cells_column_labels()) |>
    tab_style(style = cell_fill(color = "#1D3557"), locations = cells_column_spanners()) |>
    tab_style(style = cell_text(color = "white", weight = "bold"), locations = cells_column_spanners())
}
```

# Overview

## Row 1 {height="30%"}

```{r}
#| content: valuebox
#| title: "Countries Analyzed"

list(
  icon = "globe",
  color = "#D3D3D3",
  value = n_distinct(all_countries_rq2)
)
```

```{r}
#| content: valuebox
#| title: "Most Common Plastic Collected"

list(
  icon  = "recycle",
  color = "#D3D3D3",
  value = "PET"
)
```

```{r}
#| content: valuebox
#| title: "PET Drop: Low → High Biodiversity"

pet_diff <- population_plastics |>
  filter(
    parent_company == "Grand Total", grand_total > 0,
    !is.na(threatened_species), !is.na(forested_area)
    ) |>
  mutate(
    pet_prop = pet/grand_total, hdpe_prop = hdpe/grand_total,
    threatened_species_level = cut(threatened_species,
                                   breaks = c(0, 131, 556, Inf),
                                   labels = c("Low (< 131)", "Medium (131-556)", "High (> 556)"))
    ) |>
  group_by(threatened_species_level) |>
  summarize(avg_pet = mean(pet_prop, na.rm = TRUE)) |>
  summarize(pet_drop = first(avg_pet) - last(avg_pet)) |>
  pull(pet_drop)

list(
  icon  = "arrow-down",
  color = "#D3D3D3",
  value = paste0(round(pet_diff * 100, 1), " pp")
)
```

```{r}
#| content: valuebox
#| title: "Significant Predictor of PET"

list(
  icon  = "tree",
  color = "#D3D3D3",
  value = "Species"
)
```

## Row 2 {height="70%"}

### Tables {.tabset}

#### Low Coverage

```{r}
#| fig-alt: "Bar chart showing bottom 5 forest coverage countries with plastic waste breakdown"
#| label: bottom 5 countries bar chart with plastic waste breakdown

bottom5 <- plastic_ecology_summary_table |>
  mutate(across(everything(),
                ~replace_na(., 0))
         ) |>
  filter(!is.na(avg_forested)) |>
  slice_min(avg_forested, n = 5)

bottom5_forest <- ggplot(bottom5, aes(x = reorder(country, avg_forested),
                                      y = avg_forested,
                                      text = glue("PET: {round(avg_pet * 100, 2)}%
                                      HDPE: {round(avg_hdpe * 100, 2)}%
                                      LDPE: {round(avg_ldpe * 100, 2)}%
                                      PP: {round(avg_pp * 100, 2)}%
                                      PS: {round(avg_ps * 100, 2)}%
                                      PVC: {round(avg_pvc * 100, 2)}%
                                      Other: {round(avg_o * 100, 2)}%"))) +
  geom_bar(stat = "identity",
           fill = "#F3D4D5") +
  coord_flip() +
  labs(y = "Forest Coverage (%)",
       x = "",
       title = "Bottom 5 Forest Coverage Countries") +
  theme_minimal()

ggplotly(bottom5_forest, tooltip = "text")
```

Hover over each bar for plastic waste breakdown:  
PET = polyester, HDPE = high density polyethylene, LDPE = low density polyethylene, PP = polypropylene, PS = polystyrene, PVC, Other

#### Medium Coverage

```{r}
#| fig-alt: "Bar chart showing medium 5 forest coverage countries with plastic waste breakdown"
#| label: medium 5 countries bar chart with plastic waste breakdown

medium5 <- plastic_ecology_summary_table |>
  mutate(across(everything(),
                ~replace_na(., 0))
         ) |>
  filter(!is.na(avg_forested),
         avg_forested >= 19) |>
  slice_min(avg_forested, n = 5)

medium5_forest <- ggplot(medium5, aes(x = reorder(country, avg_forested),
                                      y = avg_forested,
                                      text = glue("PET: {round(avg_pet * 100, 2)}%
                                      HDPE: {round(avg_hdpe * 100, 2)}%
                                      LDPE: {round(avg_ldpe * 100, 2)}%
                                      PP: {round(avg_pp * 100, 2)}%
                                      PS: {round(avg_ps * 100, 2)}%
                                      PVC: {round(avg_pvc * 100, 2)}%
                                      Other: {round(avg_o * 100, 2)}%"))) +
  geom_bar(stat = "identity",
           fill = "#F3D4D5") +
  coord_flip() +
  labs(y = "Forest Coverage (%)",
       x = "",
       title = "Medium 5 Forest Coverage Countries") +
  theme_minimal()

ggplotly(medium5_forest, tooltip = "text")
```

Hover over each bar for plastic waste breakdown:  
PET = polyester, HDPE = high density polyethylene, LDPE = low density polyethylene, PP = polypropylene, PS = polystyrene, PVC, Other

#### High Coverage

```{r}
#| fig-alt: "Bar chart showing top 5 forest coverage countries with plastic waste breakdown"
#| label: top 5 countries bar chart with plastic waste breakdown

high5 <- plastic_ecology_summary_table |>
  mutate(across(everything(),
                ~replace_na(., 0))
         ) |>
  filter(!is.na(avg_forested),
         avg_forested >= 36) |>
  slice_max(avg_forested, n = 5)

high5_forest <- ggplot(high5, aes(x = reorder(country, avg_forested),
                                      y = avg_forested,
                                      text = glue("PET: {round(avg_pet * 100, 2)}%
                                      HDPE: {round(avg_hdpe * 100, 2)}%
                                      LDPE: {round(avg_ldpe * 100, 2)}%
                                      PP: {round(avg_pp * 100, 2)}%
                                      PS: {round(avg_ps * 100, 2)}%
                                      PVC: {round(avg_pvc * 100, 2)}%
                                      Other: {round(avg_o * 100, 2)}%"))) +
  geom_bar(stat = "identity",
           fill = "#F3D4D5") +
  coord_flip() +
  labs(y = "Forest Coverage (%)",
       x = "",
       title = "High 5 Forest Coverage Countries") +
  theme_minimal()

ggplotly(high5_forest, tooltip = "text")
```

Hover over each bar for plastic waste breakdown:  
PET = polyester, HDPE = high density polyethylene, LDPE = low density polyethylene, PP = polypropylene, PS = polystyrene, PVC, Other

### Visualizations

::: {.card}
```{r}
#| message: false
#| fig-alt: "Ridge plot showing threatened species distribution across top 3 plastic types"

population_plastics |>
  filter(parent_company == "Grand Total",
         grand_total > 0,
         !is.na(threatened_species)
         ) |>
  pivot_longer(cols = all_of(top3_plastics),
               names_to = "plastic_type",
               values_to = "count") |>
  filter(count > 0) |>
  mutate(plastic_type = factor(toupper(plastic_type),
                               levels = toupper(top3_plastics))
         ) |>
  ggplot(aes(x = threatened_species,
             y = plastic_type,
             fill = plastic_type,
             weight = count)
         ) +
  geom_density_ridges(scale = 0.8, alpha = 0.7) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_fill_manual(values = c("#E8A0A0", "#C45C5C", "#8B1A1A")) +
  labs(title = "Threatened Species Counts Across Top 3 Plastic Types",
       subtitle = "Each country weighted by quantity of plastic collected",
       x = "Threatened Species Count (log scale)", y = "") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11),
        legend.position = "none")
```
PP is collected in the most biodiverse countries, while PET dominates in nations with fewer threatened species. LDPE falls in between.
:::